Star formation in the leading arm of the Magellanic stream triggered by passage through the Galactic midplane

### 2018-07-17 ms.tex
Update on Overleaf.
Intro

\newcommand{\clustername}{\textsl{NAME}}

@@

\title{A young star cluster in the Galactic halo: \\
Recent star formation associated with the Leading Arm of the Magellanic stream}

@@

@@

We report the discovery of a young ($\lesssim 100~\textrm{Myr}$), metal-poor ($[\textrm{Fe}/\textrm{H}] \sim -1$) star cluster in the halo of the Milky Way.
The cluster --- \clustername\ --- is likely associated with the leading arm of the gas stream emanating from the Magellanic cloud system and is located $\approx 60^\circ$ from the LMC center, on the other side of the Milky Way disk relative to the LMC.
By assuming the cluster is co-located with HI gas in the stream, we, for the first time, directly measure the distance to the Magellanic stream, $d \approx 25~\textrm{kpc}$.
At this location relative to the LMC, this distance is inconsistent with predictions from models of the LMC/SMC interaction and infall into the Milky Way.
The estimated age of \clustername\ is consistent with the time of last passage through the Galactic midplane.
We therefore conclude that this star-formation event was triggered by the last disk passage, which occurred at a Galactocentric radius $R \approx XX~\kpc$.

@@

The stellar halo of the Milky Way (MW) is characterized by its old ($\approx XX$--$YY~\textrm{Gyr}$), metal-poor ($\feh \approx -1.5$) stellar population.
This is understood as a signature of the dominant (in stellar mass) progenitor systems that were accreted and disrupted early on in the MW's formation: massive dwarf galaxies \citep[e.g.,][]{Deason:2015, Fiorentino:2015}.
These systems likely came in with significant gas reservoirs, but were quenched and stripped of their gas, leading to the typically old stellar halo population in place today.
(Gas stripped now too hot/diffuse to form stars)

However, the MW continues to accrete satellite galaxies, as is evidenced by the prominent stellar stream from the Sagittarius dwarf galaxy (cite), and the presence of the Large and Small Magellanic Clouds (LMC, SMC).
While Sagittarius was likely stripped of its neutral gas long ago (cite), the LMC--SMC system contains $\approx XX~\msun$ of HI gas (cite), which form leading and trailing gas streams that cover $\approx 200^\circ$ on the sky (cite nidever).
By modeling the LMC--SMC interaction and infall, ... sims to reproduce gas velocity vs. longitude \citep{Besla:2010}
(But lacks distance constraints)
(Many people search for stars associated with the stream, but must be diffuse)

(Gas near the LMC/SMC is forming stars \citep{Casetti-Dinescu2014,MoniBidin2017}, and cite Deason/Belokurov bridge paper)

@@

Figure \ref{fig_skypm} shows the spatial and proper motion distribution in the region of interest with the objects in
a small proper motion selection shown as black and red dots.  The color magnitude diagram (CMD) of the same objects
is shown in Figure \ref{fig_cmd} and clearly reveals a sequence of objects that follow the shape of a young main-sequence.
The PS1 photometry of the same region is shown in Figure \ref{fig_isochrone} and as well as young, metal-poor isochrones at fa distance of 25 kpc.  Finally, Figure \ref{fig_gass} shows a position-velocity diagram at the position of the cluster of the \citep[GASS;][]{McClure-Griffiths2009} \hi data that has been Gaussian decomposed using the technique in \citet{Nidever2008}.
The component of the LA II can be seen on the left-hand side of the figure and a similar \hie stream can be seen near the
position of the cluster.
% Yumi:
% Need a sentence or two for dust extinction correction somewhere in the above two paragraphs?

@@

We have used the new Gaia data to detect a young, metal-poor cluster in the halo of the Milky Way at a distance of 25 kpc and a MW z-height of 12 kpc.  The detection of such a young cluster so far from the star-forming disks of any nearby
galaxy is highly unusual.  However, the proximity of the new cluster to the Leading Arm of the Magellanic Stream suggests
that the cluster originated from this metal-poor gas \citep{Fox2018}.  Previously studies have detected young stars in the Leading Arm and gas in the periphery of the Magellanic Clouds \citep{Casetti-Dinescu2014,MoniBidin2017}, however, this is the first time that an entire star cluster has been detected so far from the Clouds.

We evaluate the hypothesis that the cluster formed in the Leading Arm.  The mean proper motion values in the Magellanic Stream coordinate system \citep{Nidever2008} are ($\mu_{\rm MSL}$,$\mu_{\rm MSB}$)=($+$1.X,0.X) mas yr$^{-1}$ which is very similar to the mean values of the LMC and SMC with ($+$1.X,0.X) mas yr$^{-1}$ and ($+$1.X,0.X) mas yr$^{-1}$, respectively.  The simulations of \citet{Besla:2012} also predict proper motions at this position of approximately ($+$1.X,0.X) mas yr$^{-1}$.  The Magellanic Stream simulations give distances of the Leading Arm at this position of $\sim$50 kpc, however, many of these do not include the effects of ram pressure from the hot MW halo which could cause the gas to drop in its orbit to smaller Galactocentric distances (NEED TO DOUBLE-CHECK THIS).  In addition Figure \ref{fig_gass} shows that there is Leading Arm \hi~gas at the position of the position of the cluster with a velocity of \vlsr=70 \kmse.

>>>>

\newcommand{\clustername}{\textsl{NAME}}
\newcommand{\lmcsmc}{LMC--SMC}

@@

% \title{A young star cluster in the Galactic halo: \\
% Recent star formation associated with the Leading Arm of the Magellanic stream}
\title{Star formation in the leading arm of the Magellanic stream triggered by passage through the Galactic midplane}

@@

\author{Edward Schlafly}

\author{Douglas Finkbeiner}
% \affiliation{McWilliams Center for Cosmology, Department of Physics, Carnegie Mellon University, 5000 Forbes Avenue, Pittsburgh, PA 15213, USA}
% \affiliation{Institute of Astronomy, University of Cambridge, Madingley Road, Cambridge CB3 0HA, UK}

@@

We report the discovery of a young ($\lesssim 100~\textrm{Myr}$), metal-poor ($[\textrm{Fe}/\textrm{H}] \sim -1$) stellar association in the halo of the Milky Way ($(R, z) \sim (24, 16) kpc$).
The association --- \clustername\ --- is likely associated with the leading arm of the gas stream emanating from the Magellanic cloud system, but is located $\approx 60^\circ$ from the Large Magellanic Cloud (LMC) center, on the other side of the Milky Way disk relative to the LMC.
By assuming that the cluster is co-located with HI gas in the stream, we directly measure the distance to the Magellanic stream, $d \approx 25~\textrm{kpc}$.
At this location relative to the LMC, the measured distance is inconsistent with predictions from models of the LMC/SMC interaction and infall into the Milky Way. \todo{that don't account for ram pressure and gas interaction with MW disk}
The estimated age of \clustername\ is consistent with the time of last passage through the Galactic midplane.
We therefore conclude that this star-formation event was triggered by the last disk passage, which occurred at a Galactocentric radius $R \approx XX~\kpc$.

@@

The stellar halo of the Milky Way is characterized by its old ($\approx XX$--$YY~\textrm{Gyr}$), metal-poor ($\feh \approx -1.5$) stellar population.
This is understood as a signature of the dominant (in stellar mass) progenitor systems that were accreted and disrupted early on in the formation of the Galaxy: massive dwarf galaxies \citep[e.g.,][]{Deason:2015, Fiorentino:2015}.
These systems likely came in with significant gas reservoirs, but were quenched and stripped through collisional processes that heated and dispersed the gas \citep[e.g.,][]{Mayer:2006}, thus preventing immediate star formation in the deposited gas.

The Milky Way, however, continues to accrete satellite galaxies, as is evidenced by the prominent stellar stream from the Sagittarius dwarf galaxy \citep{Ibata:1994, Majewski:2003}, and the presence of the Large and Small Magellanic Clouds (LMC, SMC) within the virial radius of the Galaxy.
While Sagittarius was likely stripped of its neutral gas long ago \citep{Burton:1999, Tepper-Garcia:2018}, the \lmcsmc\ system contains $\approx 5\times 10^8~\msun$ of HI gas \citep{Bruns:2005}, which extends into leading and trailing gas streams that wrap nearly $\approx 200^\circ$ around the sky \citep{Putman:1998, Bruns:2005, Nidever:2010}.
Streams formed from tidally stripped material --- such as the Sagittarius or Palomar 5 \citep{Odenkirchen:2001} stellar streams, or gas streams such as the Magellanic stream (MS) --- encode information about the orbital history and future trajectory of their progenitor system \citep[e.g.,][]{Johnston:1999}.
For this reason, streams of debris are of interest for constraining the dark matter distribution around the Galaxy, as well as for providing a fossil record of the Milky Way's accretion history \citep{Johnston:1998, Bullock:2005}.

(Summarize magellanic stream stuffs)

Several groups have attempted to use the MS to constrain both the infall history of the \lmcsmc\ and the dark matter halo of the Milky Way (see recent review by \citealt{DOnghia:2016}).
Early models for the formation of the MS invoke ram pressure and tidal stripping to explain the sky distribution and extent of the MS, and typically require that the \lmcsmc\ system complete several orbits around the Galaxy in order to strip enough material \citep[e.g.,][]{Murai:1980, Gardiner:1994, Mastropietro:2005}.
However, later measurements of the proper motion of the LMC (\citealt{Kallivayalil:2006}, later confirmed by \gaia\ \citealt{van-der-Marel:2016}) and improved models for the infall dynamics \citep{Besla:2007} suggest that the \lmcsmc\ system could be on its first passage through the Galaxy \citep{Besla:2010, Besla:2012}.
In addition, the orbit of the LMC implied by the new proper motion measurement appears to be offset by $\approx 10^\circ$ from the MS \citep{Besla:2010}.
Put together, these recent revelations have led to new models for the formation of the MS that largely rely on past interactions between the LMC and SMC to preprocess the gas distribution before infall and eventual stripping by the tidal field of the Milky Way \citep{Besla:2012, Diaz:2012}

(More recently, proper motions (confirmed by Gaia) show that orbit of LMC differs from path of MS)
(Difficulty reproducing offset by invoking tidal stripping + ram pressure, interaction history important)
(But, questions remain because leading stream)

One critical difficulty in using the MS to constrain models of the \lmcsmc\ interaction and infall is the lack of distance information along the stream.
No significant over-density of stars have been found associated with the trailing portion of the MS \citep{Guhathakurta:1998}, thus leaving distance and tangential velocity information largely unknown.
The leading arm, however, ...
A small number of young OB stars have been found in the vicinity of the leading arm gas, with radial velocities consistent with having formed from the leading MS \citep{Casetti-Dinescu:2014, Zhang:2017}.
However, given their sparsity and concentration near the Galactic plane, it is difficult to unambiguously associate them with the MS: these could instead be runaway OB stars from the Milky Way disk.

Recent work has found

(Young OB stars tentatively associated with leading (Casetti-D))

However, the connection between the Magellanic clouds and the Magellanic Stream (MS)

For example, by modeling the \todo{XX} \citep{Besla:2010}, it is now understood that the \lmcsmc\ is likely on its first approach through the Galaxy, which has major implications for understanding the properties of the system itself and its effect on Galactic structure such as the bar, spiral arms, and outer disk \citep[e.g.,][]{Gomez:xxxx, Laporte:20xx}.

(But lacks distance constraints)
(Many people search for stars associated with the stream, but must be diffuse)

Gas associated with the LMC--SMC system has continued to form stars:

(Gas near the LMC/SMC is forming stars \citep{Casetti-Dinescu:2014, MoniBidin:2017}, and cite Deason/Belokurov bridge paper)

For the Magellanic Clouds, observations of tidally stripped gas or stars should help constrain the LMC--SMC interaction history and infall, and will help inform the expected response of the Milky Way to this massive accretion event \citep[e.g.,][]{Gomez:xxxx, Laporte:20xx}.

Leading arm features - ??? Disrupted satellites of the LMC?? Fox 2015 paper

@@

Figure \ref{fig_skypm} shows the spatial and proper motion distribution in the region of interest with the objects in
a small proper motion selection shown as black and red dots.  The color magnitude diagram (CMD) of the same objects
is shown in Figure \ref{fig_cmd} and clearly reveals a sequence of objects that follow the shape of a young main-sequence.
The PS1 photometry of the same region is shown in Figure \ref{fig_isochrone} and as well as young, metal-poor isochrones at fa distance of 25 kpc.  Finally, Figure \ref{fig_gass} shows a position-velocity diagram at the position of the cluster of the \citep[GASS;][]{McClure-Griffiths:2009} \hi data that has been Gaussian decomposed using the technique in \citet{Nidever:2008}.
The component of the LA II can be seen on the left-hand side of the figure and a similar \hie stream can be seen near the
position of the cluster.
% Yumi:
% Need a sentence or two for dust extinction correction somewhere in the above two paragraphs?

Isochrone: \citep{Fox:2018}

@@

We have used the new Gaia data to detect a young, metal-poor cluster in the halo of the Milky Way at a distance of 25 kpc and a MW z-height of 12 kpc.  The detection of such a young cluster so far from the star-forming disks of any nearby
galaxy is highly unusual.  However, the proximity of the new cluster to the Leading Arm of the Magellanic Stream suggests
that the cluster originated from this metal-poor gas \citep{Fox:2018}.  Previously studies have detected young stars in the Leading Arm and gas in the periphery of the Magellanic Clouds \citep{Casetti-Dinescu:2014, MoniBidin:2017}, however, this is the first time that an entire star cluster has been detected so far from the Clouds.

We evaluate the hypothesis that the cluster formed in the Leading Arm.  The mean proper motion values in the Magellanic Stream coordinate system \citep{Nidever:2008} are ($\mu_{\rm MSL}$,$\mu_{\rm MSB}$)=($+$1.X,0.X) mas yr$^{-1}$ which is very similar to the mean values of the LMC and SMC with ($+$1.X,0.X) mas yr$^{-1}$ and ($+$1.X,0.X) mas yr$^{-1}$, respectively.  The simulations of \citet{Besla:2012} also predict proper motions at this position of approximately ($+$1.X,0.X) mas yr$^{-1}$.  The Magellanic Stream simulations give distances of the Leading Arm at this position of $\sim$50 kpc, however, many of these do not include the effects of ram pressure from the hot MW halo which could cause the gas to drop in its orbit to smaller Galactocentric distances (NEED TO DOUBLE-CHECK THIS).  In addition Figure \ref{fig_gass} shows that there is Leading Arm \hi~gas at the position of the position of the cluster with a velocity of \vlsr=70 \kmse.
